### 2018-03-13 submission/manuscript.Rmd
add pat's discussion suggestions

We performed a meta-analysis to identify and validate microbiome-base biomarkers that could be used to classify individuals as having normal colons or colonic tumors using fecal or tissue samples. To our surprise, Random Forest classification models constructed to differentiate individuals with normal colons from those with carcinomas using a subset of the community performed well relative to models constructed using the full communities. When we applied the models trained on each dataset to the other datasets in our study, we found that the models trained using the subset of the communities performed better than those using the full communities. These models were trained using data in which sequences were assigned to bacterial taxa using a classifier that typically assigned sequences to the family or genus level. When we attempted to improve the specificity of the classification by using an OTU-based approach the resulting models performed as well as those constructed using coarse taxonomic assignments. These results are significant because they strengthen the growing literature indicating a role of the microbiome in tumorigenesis [@zackular_gut_2013] and as a potential tool as a non-invasive diagnostic and for assessing risk of disease and recurrence [@baxter_microbiota-based_2016; @Sze2017].

These results suggested that fine scale classification of sequences into OTUs does not improve our classification models. This has been suggested in previous literature where shotgun metagenomic data did not perform better than 16S rRNA gene sequencing data in classifying individuals with normal colons and those with carcinomas [@Hannigan2017]. We hypothesize that fine scale classification may not result in better classification because distribution of microbiota between individuals is patchy. In contrast, models using coarser taxonomic assignments will pool the fine scale diversity, resulting in less patchiness and better classification. Furthermore, the ability of models trained using a subset of the community to outperform those using the full community when testing the models on the other datasets may also be a product of the patchiness of the human-associated microbiota. The models based on the 8 taxa that had significant ORs used taxa that were found in every study and tended to have higher relative abundances. Similar to the OTU-based models, those models based on the full community taxonomy assignments were still sensitive to the patchy distribution of taxa. Regardless, it is encouraging that a collection of  `r length(crc_OR_stool$measure)` taxa could reliably classify individuals as having carcinomas considering the differences in cohorts, DNA extraction procedures, regions of the 16S rRNA gene, and sequencing methods.

When used separately to classify individuals with carcinomas, the taxa with significant ORs could not reliably classify individuals [Figure 3]. This result further supports the hypothesis that carcinoma-associated microbiota have a patchy distribution. Two individuals may have had the same classification, based on the relative abundance of different populations within this group of `r length(crc_OR_stool$measure)` taxa. Although these results only reflect associations with disease, it is tempting to hypothesize that the patchiness represents distinct mechanisms of exacerbating tumorigenesis or that multiple taxa have the same mechanism of exacerbating tumorigenesis. For example, strains of *Escherichia coli* and *Fusobacterium nucleatum* have been shown to worsen inflammation in mouse models of tumorigenesis [@ecoli_Arthur_2014; @arthur_intestinal_2012; @kostic_fusobacterium_2013]. In contrast to the patchiness of the taxa that were positively associated with carcinomas, potentially beneficial taxa had a more consistent association [Figure 6]. This result was particularly interesting because members of these taxa (i.e. *Ruminococcus* and *Clostridium XI* in stool and *Dorea* and *Blautia* in tissue) are thought to be beneficial due to their involvement in production of anti-inflammatory short chain fatty acids [@Venkataraman2016; @Herrmann2018; @Reichardt2017].

All of the adenoma classification models performed poorly and is not inconsistent with previous studies [@hale_shifts_2017; @Sze2017]. However, the classification results are at odds with results of the multitarget microbiota test (MMT) from Baxter, *et al.* [@baxter_microbiota-based_2016] who observed an AUC of 0.755 when applied to individuals with adenomas. There are two major differences between the models generated in this meta-analysis and that analysis. The MMT attempted to classify individuals as having a normal colon or having colonic lesions (i.e. adenomas or carcinomas) and not adenomas alone. Further, the MMT incorporated fecal immunoglobulin test (FIT) data while our models only used 16S rRNA gene sequencing data. Because FIT data were not available for the other studies in our meta-analysis, it was not possible to validate the MMT approach. The ability to differentiate between individuals with and without adenomas is an important problem since early detection of tumors is critical. However, it is possible that we might have been able to detect differences in the bacterial community if individuals with non-advanced and advanced adenomas were separated. This is a clinically relevant distinction since advanced adenomas are at highest risk of progressing to a carcinoma. The initial changes of the microbiota during tumorigenesis could be focal to where the initial adenoma develops and would not be easily assessed using fecal samples from an individual with non-advanced adenomas. Unfortunately, distinguishing between individuals with advanced and non-advanced adenomas was not possible in our meta-analysis since the studies did not provide the clinical data needed to make that distinction.

Stool samples represent a non-invasive approach to assessing the structure of the gut microbiota and are potentially useful for diagnosing individuals as having colonic tumors. However, they do not reflect the structure of the mucosal microbiota [@Flynn_preprint_2017]. Regardless, the taxa that were the most important in the stool-based models overlapped with those from the models trained using the data from unmatched and matched colon tissue samples [Figure S3]. Mucosal biopsies are preferred for focused mechanistic studies and have offered researchers the opportunity to sample healthy and diseased tissue from the same individuals (i.e. matched) using each individual as their own control or in a cross sectional design (i.e. unmatched). Because obtaining these samples is invasive, carries risks to the individual, and is expensive, studies investigating the structure of the mucosal microbiota generally have a limited number of participants. Thus, it was not surprising that tissue-based studies did not provide clearer associations between the mucosal microbiota and the presence of tumors. Interestingly, Fusobacterium, which has received increased recent attention for its potential role in tumorigenesis [@kostic_fusobacterium_2013] was not consistently identified across the studies in our meta-analysis. This could be due to the relatively small number of individuals in the limited number of studies. The classification models trained using the tissue-based data performed well when tested with the training data (Figure S4), but performed poorly when tested on the other tissue-associated datasets (Figure S5). Disturbingly, taxa that are commonly associated with reagent contamination (e.g. *Novosphingobium*, *Acidobacteria Gp2*, *Sphingomonas*, etc.) were detected within the tissue datasets. Such contamination is common in studies where there is relatively low bacterial biomass [@Salter_contamination_2014]. The lack of replication among the tissue-based biomarkers may be a product of the relatively small number of studies and individuals per study and possible reagent contamination.

Among our stool samples, we failed to identify several notable populations that are commonly associated with carcinomas including an enterotoxigenic strain of *Bacteroides fragilis* (ETBF) and *Streptococcus gallolyticus* subsp. *gallolyticus* [@bfrag_Boleij_2014; @strep_Aymeric_2017]. ETBF have been found in tumors in the proximal colon where they tend to form biofilms [@Purcell2017; @dejea_microbiota_2014]. Considering DNA from bacteria that are more prevalent in the proximal colon may be degraded by the time it leaves the body, it is not surprising that we failed to identify a significant OR for Bacteroides with carcinomas. In addition, since our approach could only classify sequences to the genus level and there are likely multiple Bacteroides populations in the colon, it is possible that sequences from ETBF and non-oncogenic Bacteroides were pooled. This would then reduce the OR between Bacterioides and whether an individual had carcinomas. It is also necessary to distinguish between populations that are biomarkers for a disease and those that are known to cause disease. The former may also have a causative role in the disease. Although the latter have been shown to have a causative role, they may appear at low relative abundance, be found in specific locations, or may have a highly patchy distribution among affected individuals.

Meta-analyses are a useful tool in microbiome research because they can demonstrate whether a result can be replicated and facilitate new discoveries by pooling multiple independent investigations. There have been several meta-analyses similar to this study that have sought biomarkers for obesity [@Sze2016; @Walters2014; @Finucane2014], inflammatory bowel disease [@Walters2014], and colorectal cancer [@shah_leveraging_2017]. Considering microbiome research is particularly prone to hype and overgeneralization of results, these analyses are critical. For example, previous meta-analyses have demonstrated that there are no clear fecal biomarkers for obesity [Sze2016; @Walters2014; @Finucane2014]. Such meta-analyses are difficult to perform because the underlying 16S rRNA gene sequence data are not publicly available, metadata are missing, incomplete, or vague, sequence data are of poor quality or derived by non-standard approaches, and the original studies were significantly underpowered. Reluctance to publish negative results (i.e. the “file drawer effect”) is also likely to skew our understanding of the relationship between microbiota and disease. Better attention to these specifics will increase the reproducibility and replicability of microbiota studies and make it easier to perform these crucial meta-analyses. Moving forward, meta-analyses will be important tools to help aggregate and find commonalities across studies when investigating the microbiota in the context of a specific disease [@Sze2016; @Walters2014; @Finucane2014; @shah_leveraging_2017].

>>>>

We performed a meta-analysis to identify and validate microbiota-base biomarkers that could be used to classify individuals as having normal colons or colonic tumors using fecal or tissue samples. To our surprise, Random Forest classification models constructed to differentiate individuals with normal colons from those with carcinomas using a subset of the community performed well relative to models constructed using the full communities. When we applied the models trained on each dataset to the other datasets in our study, we found that the models trained using the subset of the communities performed better than those using the full communities. These models were trained using data in which sequences were assigned to bacterial taxa using a classifier that typically assigned sequences to the family or genus level. When we attempted to improve the specificity of the classification by using an OTU-based approach the resulting models performed as well as those constructed using coarse taxonomic assignments. These results are significant because they strengthen the growing literature indicating a role for the colonic microbiota in tumorigenesis, as a potential tool as a non-invasive diagnostic, and for assessing risk of disease and recurrence [@zackular_gut_2013; @baxter_microbiota-based_2016; @Sze2017].

Fine scale classification of sequences into OTUs did not improve our classification models. This was also tested in earlier efforts to use shotgun metagenomic data to classify individuals as having normal colons or tumors; however, it was shown that analyses performed using shotgun metagenomic data did not perform better than using 16S rRNA gene sequencing data [@Hannigan2017]. We hypothesize that fine scale classification may not result in better classification because distribution of microbiota between individuals is patchy. In contrast, models using coarser taxonomic assignments will pool the fine scale diversity, resulting in less patchiness and better classification. Furthermore, the ability of models trained using a subset of the community to outperform those using the full community when testing the models on the other datasets may also be a product of the patchiness of the human-associated microbiota. The models based on the 8 taxa that had significant ORs used taxa that were found in every study and tended to have higher relative abundances. Similar to the OTU-based models, those models based on the full community taxonomy assignments were still sensitive to the patchy distribution of taxa. Regardless, it is encouraging that a collection of  `r length(crc_OR_stool$measure)` taxa could reliably classify individuals as having carcinomas considering the differences in cohorts, DNA extraction procedures, regions of the 16S rRNA gene, and sequencing methods.

When used to classify individuals with carcinomas, the taxa with significant ORs could not reliably classify individuals on their own [Figure 3]. This result further supports the hypothesis that carcinoma-associated microbiota have a patchy distribution. Two individuals may have had the same classification, based on the relative abundance of different populations within this group of `r length(crc_OR_stool$measure)` taxa. Although these results only reflect associations with disease, it is tempting to hypothesize that the patchiness is indicative of distinct mechanisms of exacerbating tumorigenesis or that multiple taxa have the same mechanism of exacerbating tumorigenesis. For example, strains of *Escherichia coli* and *Fusobacterium nucleatum* have been shown to worsen inflammation in mouse models of tumorigenesis [@ecoli_Arthur_2014; @arthur_intestinal_2012; @kostic_fusobacterium_2013]. In contrast to the patchiness of the taxa that were positively associated with carcinomas, potentially beneficial taxa had a more consistent association [Figure 6]. This result was particularly interesting because members of these taxa (i.e. *Ruminococcus* and *Clostridium XI* in fecal samples and *Dorea* and *Blautia* in tissue) are thought to be beneficial due to their involvement in production of anti-inflammatory short chain fatty acids [@Venkataraman2016; @Herrmann2018; @Reichardt2017].

All of the adenoma classification models performed poorly, which is consistent with previous studies [@hale_shifts_2017; @Sze2017]. However, the classification results are at odds with results of the multitarget microbiota test (MMT) from Baxter, et al. [@baxter_microbiota-based_2016] who observed an AUC of 0.755 when the test was applied to individuals with adenomas. There are two major differences between the models generated in this meta-analysis and that analysis. The MMT attempted to classify individuals as having a normal colon or having colonic lesions (i.e. adenomas or carcinomas) and not adenomas alone. Further, the MMT incorporated fecal immunoglobulin test (FIT) data while our models only used 16S rRNA gene sequencing data. Because FIT data were not available for the other studies in our meta-analysis, it was not possible to validate the MMT approach. The ability to differentiate between individuals with and without adenomas is an important problem since early detection of tumors is critical to patient survivorship. However, it is possible that we might have been able to detect differences in the bacterial community if individuals with non-advanced and advanced adenomas were separated. This is a clinically relevant distinction since advanced adenomas are at highest risk of progressing to carcinomas. The initial changes of the microbiota during tumorigenesis could be focal to where the initial adenoma develops and would not be easily assessed using fecal samples from an individual with non-advanced adenomas. Unfortunately, distinguishing between individuals with advanced and non-advanced adenomas was not possible in our meta-analysis since the studies did not provide the clinical data needed to make that distinction.

Fecal samples represent a non-invasive approach to assess the structure of the gut microbiota and are potentially useful for diagnosing individuals as having colonic tumors. However, they do not reflect the structure of the mucosal microbiota [@Flynn_preprint_2017]. Regardless, the taxa that were the most important in the feces-based models overlapped with those from the models trained using the data from unmatched and matched colon tissue samples [Figure S3]. Mucosal biopsies are preferred for focused mechanistic studies and have offered researchers the opportunity to sample healthy and diseased tissue from the same individuals (i.e. matched) using each individual as their own control or in a cross-sectional design (i.e. unmatched). Because obtaining these samples is invasive, carries risks to the individual, and is expensive, studies investigating the structure of the mucosal microbiota generally have a limited number of participants. Thus, it was not surprising that tissue-based studies did not provide clearer associations between the mucosal microbiota and the presence of tumors. Interestingly, *Fusobacterium*, which has received increased attention for its potential role in tumorigenesis [@kostic_fusobacterium_2013] was not consistently identified across the studies in our meta-analysis. This could be due to the relatively small number of individuals in the limited number of studies. The classification models trained using the tissue-based data performed well when tested with the training data (Figure S4), but performed poorly when tested on the other tissue-associated datasets (Figure S5). Disturbingly, taxa that are commonly associated with reagent contamination (e.g. *Novosphingobium*, *Acidobacteria Gp2*, *Sphingomonas*, etc.) were detected within the tissue datasets. Such contamination is common in studies where there is relatively low bacterial biomass [@Salter_contamination_2014]. The lack of replication among the tissue-based biomarkers may be a product of the relatively small number of studies and individuals per study and possible reagent contamination.

Among the fecal sample data, we failed to identify several notable populations that are commonly associated with carcinomas including an enterotoxigenic strain of *Bacteroides fragilis* (ETBF) and *Streptococcus gallolyticus* subsp. *gallolyticus* [@bfrag_Boleij_2014; @strep_Aymeric_2017]. ETBF have been found in tumors in the proximal colon where they tend to form biofilms [@Purcell2017; @dejea_microbiota_2014]. Considering DNA from bacteria that are more prevalent in the proximal colon may be degraded by the time it leaves the body, it is not surprising that we failed to identify a significant OR for *Bacteroides* with carcinomas. In addition, since our approach could only classify sequences to the genus level and there are likely multiple *Bacteroides* populations in the colon, it is possible that sequences from ETBF and non-oncogenic *Bacteroides* were pooled. This would then reduce the OR between *Bacterioides* and whether an individual had carcinomas. It is also necessary to distinguish between populations that are biomarkers for a disease and those that are known to cause disease. Although the latter have been shown to have a causative role, they may appear at low relative abundance, be found in specific locations, or may have a highly patchy distribution among affected individuals.

Meta-analyses are a useful tool in microbiome research because they can demonstrate whether a result can be replicated and facilitate new discoveries by pooling multiple independent investigations. There have been several meta-analyses similar to this study that have sought biomarkers for obesity [@Sze2016; @Walters2014; @Finucane2014], inflammatory bowel disease [@Walters2014], and colorectal cancer [@shah_leveraging_2017]. Considering microbiome research is particularly prone to hype and overgeneralization of results [@Hanage2014], these analyses are critical. Meta-analyses are difficult to perform because the underlying 16S rRNA gene sequence data are not publicly available, metadata are missing, incomplete, or vague, sequence data are of poor quality or derived by non-standard approaches, and the original studies may be significantly underpowered. Reluctance to publish negative results (i.e. the “file drawer effect”) is also likely to skew our understanding of the relationship between microbiota and disease. Better attention to these specific issues will increase the reproducibility and replicability of microbiota studies and make it easier to perform these crucial meta-analyses. Moving forward, meta-analyses will be important tools to help aggregate and find commonalities across studies when investigating the microbiota in the context of a specific disease [@Sze2016; @Walters2014; @Finucane2014; @shah_leveraging_2017].
